Spitzer, Gaia and the Potential of the Milky Way

### 2013-07-17 text/paper1/paper.tex
text changes
some updates from KVJ comments

\author{Johnston, Price-Whelan, Madore, Majewski}

\begin{abstract}
sdfsdf
\end{abstract}

@@

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide us with a stellar distance
indicator of unprecedented accuracy and precision. \citet{madore2012}
have demonstrated \citep[using 5 stars with trigonometric parallaxes
  measured by Hubble;][]{benedict2011} that the dispersion in the
mid-IR PL relation -- at Spitzer wavelengths -- is of order
$\pm$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $\pm 2\%$ for RR
Lyrae stars out to $\sim$80~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables, for modest integration
times). [Distance measurements of Blue Horizontal Branch and Blue Straggler stars in the halo typically achieve 15-25\% uncertainties.]

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia2001}, planned for launch in October
2013, is an astrometric mission which, amongst a myriad other
science goals, aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
over the 5-year lifetime of the mission (cite?), Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full six-dimensional phase-space
maps of the Galaxy for heliocentric distances of up to $\sim$ 10~kpc (xx kpc)
for $V\sim xxx$ ($V\sim yyy$) stars.

@@

The anticipated proper motion errors for RR Lyrae correspond to
tangential velocity errors of order $\sim$1~km/s at 30~kpc, and of
order $\sim$10~km/s out to several tens of kiloparsecs. The
combination of Spitzer and Gaia data will extend the `horizon' of
where full six-dimensional phase-space maps of the Galaxy are possible
from $<10$~kpc to 80~kpc. This enormous increase in volume will refine
data on existing debris systems such as Sagittarius, the Orphan
stream, and [...], and enable discovery of new substructure within the
Galactic halo.

@@

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} illustrates results from the LM10
N-body simulation of dwarf satellite disruption along the expected Sgr
orbit. The particle positions are shown in a coordinate system defined
by the orbit of the satellite orbit \citep{majewski2004}. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2).
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ flattening ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk:
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right)
\end{align}

@@

The combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris opens up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc is smaller than
the predicted thickness of the stream, ranging from
$\sim$2-$10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\sim$2~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream \citep[$\sigma_v\sim$10~km/s;][]{majewski2004}, and ground-based radial
velocities with $\sim$10~km/s accuracy, the measured phase-space
position of each individual RR Lyrae relative to the stream becomes a
powerful probe of the potential: the next section outlines a new
method to take advantage of this information.

\section{Method}
\label{sec:method}
With access to full phase-space information for stars in a tidal
stream, it is no longer necessary to model the orbit of the stream
itself. Instead, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day orbital parameters of the
progenitor system are known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time, the
stars should `recombine' with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system.

Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_i, \bs{v}_i)|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi$.\footnote{Future work will
  investigate a more general ... basis function blah blah.} For a
given set of potential parameters, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates, $(\bs{q}_i(t_j), \bs{p}_i(t_j))$, are
computed using the instantaneous tidal radius, escape velocity, and
phase-space position of the progenitor. At a particular timestep,
these coordinates are defined as
\begin{equation}
  \bs{q}_i = \frac{\bs{x}_i -
    \bs{x}_p}{R_{tide}}\,\,\,\,\,\,\,\,\bs{p}_i = \frac{\bs{v}_i -
    \bs{v}_p}{v_{esc}}
\end{equation}
where $\bs{x}_i$ and $\bs{v}_i$ are the 3D position and velocity of
the $i$th particle, and $\bs{x}_p$ and $\bs{v}_p$ are the same for the
progenitor.  These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the computed enclosed mass
of the potential, $m_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{tide}=\Big(\frac{m_{sat}}{3m_{enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{esc}=\sqrt{\frac{2Gm_{sat}}{R_{tide}}}.
\end{equation}

Qualitatively, when the distance in this normalized 6D space,
$D_{ps}=\sqrt{|\bs{q}_i|^2+|\bs{p}_i|^2} < \sqrt{2}$, the star is
probably recaptured by the satellite, but this is not imposed as a
hard constraint.  As a statistical measure of how many stars have recombined, we
construct the 6D distribution of the minimum phase-space vectors for
each star over each individual, integrated orbit. For each star, $i$,
the phase-space distance, $D_{ps}$, is computed at each timestep
$t_j$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_i &= \argmin_{t} D_{ps,i}\\
  A_{ik} &= (\bs{q}_i(t^*_i),\bs{p}_i(t^*_i)) % A_ik are really samples from a distribution, or we treat them as such
\end{align}
where $k\in\{1,6\}$. Intuitively, the variance of this distribution
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars; in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik})).
\end{equation}
[This objective function is an approximation to a more general likelihood function, described in detail in forthcoming work (Price-Whelan \& Johnston in prep.). The results shown in sections xx and yy are meant to illustrate the strength and concept of the method..]

\subsection{Application to Simulated Data}
The LM10 (see Section \ref{sec:sgr}) simulation data is a perfect
test-bed for evaluating the effectiveness of this method. We start by
extracting both particle data and the satellite orbital parameters
from the present-day snapshot of the simulation
data.\footnote{\url{www.astro.virginia.edu/~srm4n/Sgr/data.html}} We then
`observe' a sample of 100 stars from the first leading and trailing
wraps of the stream: the radial velocity errors are drawn from a
Gaussian $\epsilon_{rv} \sim \mathcal{N}(\mu=0,\sigma=10~{\rm km/s})$,
the tangential velocity errors depend on the distance to the star and
are computed from the expected Gaia error curve (shown in
Figure~\ref{fig:gaia}), and the distance errors on a star of distance
$D$ are $\epsilon_{dist} \sim \mathcal{N}(0,0.02\times
D)$. Figure~\ref{fig:} shows all 2D projections of the 6D minimum
phase-space distance distribution for the observed particles in the
correct and an incorrect potential. Qualitatively, it is clear that the
variance of the distribution increases in the wrong potential. The objective function
(generalized variance) defines a convex hypersurface over which we
optimize four of the six logarithmic potential parameters:
$v_{circ}$, $\phi$, $q_1$, and $q_z$. Figure~\ref{fig:} shows [...]
curves produced by varying the true parameters of the potential by
10\% around the true value.

% Discussion?
Many orbit fitting methods for measuring the Galactic potential assume that the stars in a given stream are on the same
orbit. \cite{binney??} show that this assumption leads to biased
potential parameters, even without realistic observational errors on
the kinematic information. The method presented above is
free of this assumption and instead \emph{relies} on the fact that
each star is on a different orbit and is thus an independent potentiometer. The power of this method is in
the joint constraint that the best potential is the one that gets the
most of these individual orbits to intersect the progenitor
over the history of interaction with the system.

@@

\subsection{Future prospects}

@@

>>>>

\author{Adrian~M.~Price-Whelan\altaffilmark{1,2} Kathryn V. Johnston\altaffilmark{1}}

% affiliations
\altaffiltext{1}{Department of Astronomy, Columbia University, 550 W 120th St., New York, NY 10027, USA}
\altaffiltext{2}{NSF Graduate Fellow}

\begin{abstract}
.
\end{abstract}

@@

These studies suggest that observations of RR Lyrae variables by
NASA's Spitzer mission could provide us with a stellar distance
indicator of unprecedented accuracy and precision. \citet{madore2012}
have demonstrated \citep[using 5 stars with trigonometric parallaxes
  measured by Hubble;][]{benedict2011} that the dispersion in the
mid-IR PL relation -- at Spitzer wavelengths -- is of order
$\pm$0.03~mag. This implies that it is possible to determine
individual distances that are good to better than $\pm 2\%$ for RR
Lyrae stars out to $\sim$80~kpc (the limit of Spitzer's ability to
detect and measure RR Lyrae variables, for modest integration
times). Distance measurements of Blue Horizontal Branch and Blue Straggler stars at these distances in the halo typically achieve 15-25\% uncertainties. [can cite Alis Deason if necessary].

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia2001}, planned for launch in October
2013, is an astrometric mission which aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
over the 5-year lifetime of the mission, Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full six-dimensional phase-space
maps of the Galaxy with $<$10\% distance errors for heliocentric distances of up to $\sim$10~kpc for ...

@@

The anticipated proper motion errors for RR Lyrae correspond to
tangential velocity errors of order $\sim$1~km/s at 30~kpc, and of
order $\sim$10~km/s out to several tens of kiloparsecs. The
combination of Spitzer and Gaia data will extend the `horizon' of
where full six-dimensional phase-space maps of the Galaxy are possible
from $<10$~kpc to 80~kpc. This enormous increase in volume will refine
data on existing debris systems such as the Sagittarius stream.

@@

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} illustrates results from the LM10
N-body simulation of dwarf satellite disruption along the expected Sgr
orbit. The particle positions are shown in a coordinate system defined
by the orbit of the satellite orbit \citep{majewski2004}. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2).
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ flattening ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right)
\end{align}

@@

The combined Spitzer and Gaia measurements of distances and proper
motions of RR Lyraes in the Sgr debris opens up new avenues of
interpretation. A 2\% distance error at $\sim$30~kpc is smaller than
the predicted thickness of the stream, ranging from
$\sim$2-$10$~kpc. Combined with Gaia tangential velocity measurement
accuracies ($\sim$2~km/s at 30~kpc) less than the intrinsic dispersion
for much of the stream \citep[$\sigma_v\sim$10~km/s;][]{majewski2004}, and ground-based radial
velocities with $\sim$10~km/s accuracy, the measured phase-space
position of each individual RR Lyrae relative to the stream becomes a
powerful probe of the potential (see \ref{fig:}): the next section outlines a new
method to take advantage of this information.

\section{Description and test of our algorithm}
\label{sec:method}
With access to full phase-space information for stars in a tidal
stream, it is no longer necessary to model the orbit of the stream
itself. Instead, it is possible to use the individual stars as
potential measures by exploting the fact that the stars must have come
from the same progenitor. If the present-day orbital parameters of the
progenitor system are known, this places an intuitive constraint on
the potential of the Milky Way: if the orbits of the stars and the
orbit of the progenitor are integrated \emph{backwards} in time, the
stars should `recombine' with the progenitor. If the potential is
incorrect, the orbits of the stars will diverge from that of the
progenitor and thus will not be recaptured by the satellite system.

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi$. For a
given set of potential parameters, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the computed enclosed mass
of the potential, $m_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=\Big(\frac{m_{\rm sat}}{3m_{\rm enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}

Qualitatively, when the distance in this normalized 6D space,
$D_{\rm ps}=\sqrt{|\bs{q}_{i}|^2+|\bs{p}_{i}|^2} < \sqrt{2}$, the star is
probably recaptured by the satellite, but this is not imposed as a
hard constraint. [JZSH ...] As a statistical measure of how many stars have recombined, we
construct the 6D distribution of the minimum phase-space vectors for
each star over each individual, integrated orbit. For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{\rm ps,i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in\{1,6\}$. Intuitively, the variance of this distribution
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars; in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik})).
\end{equation}
[This objective function is an approximation to a more general likelihood function, described in detail in forthcoming work (Price-Whelan \& Johnston in prep.). The results shown in sections xx and yy are meant to illustrate the strength and concept of the method..]

\subsection{Application to Simulated Data}
The LM10 simulation data (see Section \ref{sec:sgr}) is a perfect
test-bed for evaluating the effectiveness of this method. We start by
extracting both particle data and the satellite orbital parameters
from the present-day snapshot of the simulation
data.\footnote{\url{www.astro.virginia.edu/~srm4n/Sgr/data.html}} We then
`observe' a sample of 100 stars from the first leading and trailing
wraps of the stream: the radial velocity errors are drawn from a
Gaussian $\epsilon_{rv} \sim \mathcal{N}(\mu=0,\sigma=10~{\rm km/s})$,
the tangential velocity errors depend on the distance to the star and
are computed from the expected Gaia error curve (shown in
Figure~\ref{fig:gaia}), and the distance errors on a star of distance
$D$ are $\epsilon_{dist} \sim \mathcal{N}(0,0.02\times
D)$. Figure~\ref{fig:} shows all 2D projections of the 6D minimum
phase-space distance distribution for the observed particles in the
correct and an incorrect potential. Qualitatively, it is clear that the
variance of the distribution increases in the wrong potential. The objective function
(generalized variance) defines a convex hypersurface over which we
optimize four of the six logarithmic potential parameters:
$v_{circ}$, $\phi$, $q_1$, and $q_z$. Figure~\ref{fig:} shows [...]
curves produced by varying the true parameters of the potential by
10\% around the true value.

@@

\section{Discussion and future prospects}

Many orbit fitting methods for measuring the Galactic potential assume that the stars in a given stream are on the same
orbit. \cite{binney??} show that this assumption leads to biased
potential parameters, even without realistic observational errors on
the kinematic data. The method presented above is
free of this assumption and instead \emph{relies} on the fact that
each star is on a different orbit and is thus an independent potentiometer. The power of this method is in
the joint constraint that the best potential is the one that gets the
most of these individual orbits to intersect the progenitor
over the history of interaction with the system.

[merge these paragraphs (above, below)]

@@

\acknowledgments
We thank ...

APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ 11-44155.
